---
title: "Rsession_03_add_kgen_exac_filter_common_germ"
author: "Mae Goldgraben"
date: "28May2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

last updated: Mae Goldgraben, 28May2018  

# Summary

The script performs the following:  

- check the data   

- remove variants uniform across all samples (especially after GQ, DP, CR filtering)  

- incorporate exac and kgen for selected variants where available (biallelic have been selected in exac/kgen)  

- filter out common variants 5% and 1% AFs from EUR kgen  

- save data  

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Clean up
rm(list=ls())

# Load libraries
library(knitr)
library(dplyr)
library(ggplot2)

# Folders
setwd("/rds/project/erf33/rds-erf33-medgen/users/mae/IHCAP75_atm_aml/germline_analysis_all_timepts/Rscripts_and_filters")
source_data_folder <- "/rds/project/erf33/rds-erf33-medgen/users/mae/IHCAP75_atm_aml/germline_analysis_all_timepts/Rscripts_and_filters/source_data"
interim_data_folder <- "/rds/project/erf33/rds-erf33-medgen/users/mae/IHCAP75_atm_aml/germline_analysis_all_timepts/Rscripts_and_filters/interim_data"

```

# load_data

```{r load_data}

load(paste(interim_data_folder, "Rsession_02_filter_genotypes_and_variants_germ.RData", sep="/"))

```

# check_data

```{r check_data}
ls()

dim(cleaned_gt.mx)
class(cleaned_gt.mx)
cleaned_gt.mx[1:5,1:5]

dim(cleaned_vv.df)
class(cleaned_vv.df)
cleaned_vv.df[1:5,1:5]

dim(cleaned_kgen.df)
class(cleaned_kgen.df)
cleaned_kgen.df[1:5,1:5]

dim(cleaned_exac.df)
class(cleaned_exac.df)
cleaned_exac.df[1:5,1:5]

# Check consistency of rownames
sum(rownames(cleaned_gt.mx) != rownames(cleaned_vv.df))
sum(rownames(cleaned_gt.mx) != rownames(cleaned_kgen.df))
sum(rownames(cleaned_gt.mx) != rownames(cleaned_exac.df))

```

# remove_variants_with_the_uniform_genotypes_across_all_samples

Remove 8,464 (624 all ref; 2,546 all het; 5,294 all homo) variants of 55,579 -> 47,115

```{r remove_variants_with_the_uniform_genotypes_across_all_samples}

# Check that there are none all-NA variants
non_NA_count.udf <- function(x){sum(!is.na(x))}
all_NA <- apply(cleaned_gt.mx, 1, non_NA_count.udf) == 0
sum(all_NA) # 0

# Function to detect uniform numeric vector
uniform_vector.udf <- function(x){
  if(min(x, na.rm=TRUE) == max(x, na.rm=TRUE)){return(TRUE)} else {return(FALSE)}}

# Variants with uniform genotypes across all samples 
uniform_genotypes <- apply(cleaned_gt.mx, 1, uniform_vector.udf)
summary(uniform_genotypes) # 8,464
uni.gen.df <- as.data.frame(uniform_genotypes)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~#
uniform_vector.0.udf <- function(x){
  if(min(x, na.rm=TRUE) == 0 & max(x, na.rm=TRUE) == 0){return(TRUE)} else {return(FALSE)}}
uniform_genotypes0 <- apply(cleaned_gt.mx, 1, uniform_vector.0.udf)
summary(uniform_genotypes0) # 624
uni.gen.0.df <- as.data.frame(uniform_genotypes0)

uniform_vector.1.udf <- function(x){
  if(min(x, na.rm=TRUE) == 1 & max(x, na.rm=TRUE) == 1){return(TRUE)} else {return(FALSE)}}
uniform_genotypes1 <- apply(cleaned_gt.mx, 1, uniform_vector.1.udf)
summary(uniform_genotypes1) # 2,546
uni.gen.1.df <- as.data.frame(uniform_genotypes1)

uniform_vector.2.udf <- function(x){
  if(min(x, na.rm=TRUE) == 2 & max(x, na.rm=TRUE) == 2){return(TRUE)} else {return(FALSE)}}
uniform_genotypes2 <- apply(cleaned_gt.mx, 1, uniform_vector.2.udf)
summary(uniform_genotypes2) # 5,294
uni.gen.2.df <- as.data.frame(uniform_genotypes2)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~#

# Remove variants with uniform genotypes across all samples
cleaned_gt.mx <- cleaned_gt.mx[!uniform_genotypes,]
cleaned_vv.df <- cleaned_vv.df[!uniform_genotypes,]
cleaned_kgen.df <- cleaned_kgen.df[!uniform_genotypes,]
cleaned_exac.df <- cleaned_exac.df[!uniform_genotypes,]

dim(cleaned_gt.mx) # 47,115 x 8
dim(cleaned_vv.df)
dim(cleaned_kgen.df)
dim(cleaned_exac.df)

# Clean-up
rm(non_NA_count.udf, all_NA, uniform_vector.udf, uniform_genotypes, uni.gen.df, uniform_vector.0.udf, uniform_genotypes0, uni.gen.0.df, uniform_vector.1.udf, uniform_genotypes1, uni.gen.1.df, uniform_vector.2.udf, uniform_genotypes2, uni.gen.2.df)

```

# add_exac_and_kgen_frequencies_to_vv_file

```{r add_exac_kgen_AFs_to_vv_file}

# Since ALL samples in analysis were mainly EUR decent
# Add kgen overall and EURopean AFs
sum(rownames(cleaned_vv.df)!= rownames(cleaned_kgen.df))
cleaned_vv.df$kgen.AF <- cleaned_kgen.df$kgen.AF
cleaned_vv.df$kgen.EUR_AF <- cleaned_kgen.df$kgen.EUR_AF

# Add exac overall and EURopean AFs
sum(rownames(cleaned_vv.df)!= rownames(cleaned_exac.df))
cleaned_vv.df$exac_non_TCGA.AF <- cleaned_exac.df$exac_non_TCGA.AF
cleaned_vv.df$exac_non_TCGA_NFE.AF <- cleaned_exac.df$exac_non_TCGA.AC_NFE/cleaned_exac.df$exac_non_TCGA.AN_NFE

```

# filter_out_most_common_variants_EUR_kgen_in_vv_file_0.05_0.95

Keep vars with AFs in EUR kgen MAF<0.05, MAF>0.95, or NA -> 10,284 vars

```{r filter_out_most_common_variants_EUR_kgen_in_vv_file_0.05_0.95}

# pull variants kgen AF < 0.05, > 0.95, or NA
cleaned_vv_rare1.df <- cleaned_vv.df %>% filter(kgen.EUR_AF < 0.05 | kgen.EUR_AF > 0.95 | is.na(kgen.EUR_AF))
dim (cleaned_vv_rare1.df) # 10,284 x 72

rare_vars <- as.vector(cleaned_vv_rare1.df$SplitVarID)
cleaned_gt_rare1.df <- cleaned_gt.mx[rare_vars, ]
dim (cleaned_gt_rare1.df) # 10,284 x 8

# cross-check - create vector of common variants in ALL_kgen
kgen_common <- row.names(cleaned_kgen.df[cleaned_kgen.df$kgen.EUR_AF >= 0.05 & cleaned_kgen.df$kgen.EUR_AF <= 0.95 & !is.na(cleaned_kgen.df$kgen.EUR_AF),])
# Number of variants to remove
dim(cleaned_vv.df) # 47,115 x 72
length(kgen_common) # 36,831

# Remove common variants from vv file 
cleaned_vv_rare.df <- cleaned_vv.df[!row.names(cleaned_vv.df) %in% kgen_common,]
# Number of variants remaining after rare filters
nrow(cleaned_vv_rare.df) # 10,284
# Sort vv_rare file on rownames
cleaned_vv_rare.df <- cleaned_vv_rare.df[order(row.names(cleaned_vv_rare.df)),]

# Check and remove common variants from gt file 
dim(cleaned_gt.mx) # 47,115 x 8
cleaned_gt_rare.mx <- cleaned_gt.mx[!row.names(cleaned_gt.mx) %in% kgen_common,]
# Number of variants remaining after rare filters
nrow(cleaned_gt_rare.mx) # 10,284
# Sort vv_rare file on rownames
cleaned_gt_rare.mx <- cleaned_gt_rare.mx[order(row.names(cleaned_gt_rare.mx)),]

sum(rownames(cleaned_gt_rare.mx) != rownames(cleaned_gt_rare1.df)) # effectively the same!

# Clean up
rm(rare_vars, kgen_common, cleaned_vv_rare1.df, cleaned_gt_rare1.df)

```

# filter_out_all_common_variants_EUR_kgen_in_vv_file_0.01_0.99

Keep vars with AFs in EUR kgen MAF<0.01, MAF>0.99, or NA -> 5,993 vars

```{r filter_out_most_common_variants_EUR_kgen_in_vv_file_0.01_0.99}

# pull variants kgen AF < 0.01, > 0.99, or NA
cleaned_vv_rarest1.df <- cleaned_vv.df %>% filter(kgen.EUR_AF < 0.01 | kgen.EUR_AF > 0.99 | is.na(kgen.EUR_AF))
dim (cleaned_vv_rarest1.df) # 5,993 x 72

rarest_vars <- as.vector(cleaned_vv_rarest1.df$SplitVarID)
cleaned_gt_rarest1.df <- cleaned_gt.mx[rarest_vars, ]
dim (cleaned_gt_rarest1.df) # 5,993 x 8

# cross-check - create vector of common variants in EUR_kgen
kgen_commonest <- row.names(cleaned_kgen.df[cleaned_kgen.df$kgen.EUR_AF >= 0.01 & cleaned_kgen.df$kgen.EUR_AF <= 0.99 & !is.na(cleaned_kgen.df$kgen.EUR_AF),])
# Number of variants to remove
dim(cleaned_vv.df) # 47,115 x 72
length(kgen_commonest) # 41,122

# Remove common variants from vv file 
cleaned_vv_rarest.df <- cleaned_vv.df[!row.names(cleaned_vv.df) %in% kgen_commonest,]
# Number of variants remaining after rare filters
nrow(cleaned_vv_rarest.df) # 5,993
# Sort vv_rare file on rownames
cleaned_vv_rarest.df <- cleaned_vv_rarest.df[order(row.names(cleaned_vv_rarest.df)),]

# Check and remove common variants from gt file 
dim(cleaned_gt.mx) # 47,115
cleaned_gt_rarest.mx <- cleaned_gt.mx[!row.names(cleaned_gt.mx) %in% kgen_commonest,]
# Number of variants remaining after rare filters
nrow(cleaned_gt_rarest.mx) # 5,993
# Sort vv_rare file on rownames
cleaned_gt_rarest.mx <- cleaned_gt_rarest.mx[order(row.names(cleaned_gt_rarest.mx)),]

sum(rownames(cleaned_gt_rarest.mx) != rownames(cleaned_gt_rarest1.df)) # effectively the same!

# Clean up
rm(rarest_vars, kgen_commonest, cleaned_vv_rarest1.df, cleaned_gt_rarest1.df)

```

# clean-up

```{r}

rm(cleaned_kgen.df, cleaned_exac.df, cleaned_gt.mx, cleaned_vv.df)

```

# data_summary

```{r data_summary}

ls()

dim(cleaned_gt_rare.mx)
class(cleaned_gt_rare.mx)
cleaned_gt_rare.mx[1:5,1:5]

dim(cleaned_vv_rare.df)
class(cleaned_vv_rare.df)
cleaned_vv_rare.df[1:5,1:5]

dim(cleaned_gt_rarest.mx)
class(cleaned_gt_rarest.mx)
cleaned_gt_rarest.mx[1:5,1:5]

dim(cleaned_vv_rarest.df)
class(cleaned_vv_rarest.df)
cleaned_vv_rarest.df[1:5,1:5]

# Check consistency of colnames and rownames
sum(rownames(cleaned_gt_rare.mx) != rownames(cleaned_vv_rare.df))
sum(rownames(cleaned_gt_rarest.mx) != rownames(cleaned_vv_rarest.df))

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "Rsession_03_add_kgen_exac_filter_common_germ.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
